---
title: "Midterm"
author: "Yiping Li"
output: github_document
date: "`r Sys.Date()`" 
always_allow_html: true
---
Requirement: formulate a clear and concise question to answer and conduct data wrangling, exploratory data analysis, and data visualization to explore/answer this question.

```{r, packages}
library(readxl)
library(plyr)
library(dplyr)

```

Introduction: 
Interested in the association between UPFs consumption and type 2 diabetes among young adults
(related variables)

Method (put down step by step, plots): 
```{r, import raw data}
metaair <- readRDS("/Users/yiping/Desktop/Epi research/OneDrive - University of Southern California/MetaChem/metaair.rds")
UPFMean_grams <- read_excel("/Users/yiping/Desktop/Epi research/USC-Liz/independent research - UPF/UPF% calculation2/UPFgrams_baseline.xlsx")

```

```{r, data wrangling: extract variables and make a subset}
myvars <- c("ID", "cage", "male", "race2", "diabetesbinary",
            "gluc_fasting", "gluc_120min", "hba1c", "HEI", "MDS", "DII", "DASH")

subsetmetachem <- metaair[myvars]

```

```{r, data wrangling: change names of categorical varialbes into numbers}
subsetmetachem$race2 <- revalue(subsetmetachem$race2, c("White"="1", "Hispanic"="2", "Other"="3"))

subsetmetachem$diabetesbinary <- revalue(subsetmetachem$diabetesbinary, c("no diabetes"="0", "T2D or prediabetes"="1"))

```

```{r, data wrangling: merge datasets}
vars <- merge(subsetmetachem, UPFMean_grams)

```

```{r, data wrangling: make nnumeric variables into categorical variables}
vars$UPFcat <- cut(vars$UPFpercentage_mean,
                        breaks=c(0, 75, 100),
                        labels=c("0", "1"))

vars$UPFquart <- cut(vars$UPFpercentage_mean,
                          breaks=c(0, 25, 50, 75, 100),
                          labels=c("1", "2", "3", "4"))

```

```{r, data wrangling: rename column names}
names(vars)[2] <- "Age"
names(vars)[3] <- "Male"
names(vars)[4] <- "Race"
names(vars)[5] <- "DiabetesCat"
names(vars)[6] <- "Gluc_Fasting"
names(vars)[7] <- "Gluc_120min"
names(vars)[8] <- "HbA1C"

```

EDA: 
```{r, check for missing values, remove age <18, T2DM NA variables}
#skimr package

vars <- vars %>%
  filter(!is.na(DiabetesCat))

vars <- vars %>%
  filter(Age > 18)

```

Data Visualization: 
```{r, histogram, glu-fasting/120mins by UPFquart}

```

```{r, histogram, glu_fasting/120mins by UPFcat}

```

```{r, barchart, glu_fasting/120mins by UPFcat+sex --> 4 cats}

```

```{r, barchart, glu_fasting/120mins by UPFcat+race --> 6 cats}

```

```{r, statistical summary graph, }

```

```{r, summary table should include sample size}

```

Conclusion:
